Adaptive global optimization based on nested dimensionality reduction\thanks{This study was supported by the Russian Science Foundation, project No.\,16-11-10150.}

### 2019-03-19 2019/WCGO/WCGO.tex
Final

The use of Peano curve $y(x)$ 
\begin{equation}
\left\{y\in R^N: -2^{-1}\leq y_i \leq 2^{-1}, 1 \leq i \leq N\right\}=\left\{
y(x):0\leq x \leq 1 \right\}
\end{equation}
unambiguously mapping the interval of real axis $[0,1]$ onto a 
$N$-dimensional cube is the first of the dimension reduction methods considered.
Problems of numerical construction of Peano-type space-filling curves and the 
corresponding theory are considered in detail in 
\cite{Sergeyev2013,Strongin2000}. Here we will note that a numerically 
constructed curve (\textit{evolvent}) is $2^{-m}$ accurate approximation of 
the theoretical Peano curve, where $m$ is an evolvent construction parameter.

@@

The first series of experiments has been carried out on the two-dimensional problems from the classes $F_{gr}$, GKLS 
\textit{Simple}, and GKLS \textit{Hard} (100 functions from each class). Table~\ref{tab1} presents the averaged 
numbers of trials executed by GSA with the use of evolvents ($K_e$), nested optimization scheme ($K_n$), and adaptive 
nested optimization scheme ($K_a$). Fig.~\ref{fig1} and Fig.~\ref{fig2}(a,b) present the operating characteristics of the 
algorithms obtained on the problem classes $F_{gr}$, GKLS \textit{Simple}, and GKLS \textit{Hard} respectively. The solid 
line corresponds to the algorithm using the evolvents, short dashed line -- to the adaptive nested optimization scheme, and 
the long dashed line -- to the nested optimization scheme. The results of experiments demonstrate that GSA with the use of 
the adaptive nested optimization scheme shows almost the same speed as compared to GSA with the evolvents, and 
both algorithms considerably exceed the algorithm using the nested optimization scheme. Therefore further experiments 
were limited to the comparison of different variants of the adaptive  dimensionality reduction scheme.

\begin{table}
\centering
\caption{Average number of trials for 2D problems.}\label{tab1}
\begin{tabular}{lccc}
\hline\noalign{\smallskip}
 &  $\;\;\;F_{gr}\;\;\;$  &  GKLS \textit{Simple} &  GKLS \textit{Hard} \\
\noalign{\smallskip}\hline\noalign{\smallskip}
 $K_e$ & 180  & 252 & 674 \\
 $K_n$ & 341  & 697 & 1252 \\
 $K_a$ & 215  & 279 & 815 \\
\noalign{\smallskip}\hline
\end{tabular}
\end{table}

@@

The second series of experiments has been carried out on the four-dimensional problems from the classes GKLS 
\textit{Simple} and GKLS \textit{Hard} (100 functions of each class). Table~\ref{tab2}  presents the averaged numbers 
of trials executed by GSA with the use of the adaptive nested optimization scheme ($K_a$) and block adaptive nested 
optimization scheme ($K_{ba}$) with two levels of subproblems of equal dimensionality $N_1=N_2=2$. 
Note that when solving the problem of the dimensionality $N=4$ using the initial variant of the adaptive scheme, four 
levels of one-dimensional subproblems are formed that complicates the processing of these ones.

Fig.~\ref{fig3}(a,b) presents the operating characteristics of the algorithms obtained on the classes GKLS \textit{Simple} 
and GKLS \textit{Hard} respectively. The dashed line corresponds to GSA using the adaptive nested optimization 
scheme, the solid line -- the block adaptive nested optimization scheme. 
The results of experiments demonstrate the use of the block adaptive nested optimization scheme provides a considerable 
gain in the number of trials (up to 35\%) as compared to the initial adaptive nested optimization scheme.

\begin{table}
\centering
\caption{Average number of trials for 4D problems.}\label{tab2}
\begin{tabular}{lcc}
\hline\noalign{\smallskip}
 &    GKLS \textit{Simple} &  GKLS \textit{Hard} \\
\noalign{\smallskip}\hline\noalign{\smallskip}
 $K_a$  &  21747 & 35633 \\
 $K_{ba}$ &  13894 & 31620 \\
\noalign{\smallskip}\hline
\end{tabular}
\end{table}

>>>>

The use of Peano curve $y(x)$ 
\begin{equation}
\left\{y\in R^N: -2^{-1}\leq y_i \leq 2^{-1}, 1 \leq i \leq N\right\}=\left\{
y(x):0\leq x \leq 1 \right\}
\end{equation}
unambiguously mapping the interval of real axis $[0,1]$ onto an 
$N$-dimensional cube is the first of the dimension reduction methods considered.
Problems of numerical construction of Peano-type space-filling curves and the 
corresponding theory are considered in detail in 
\cite{Sergeyev2013,Strongin2000}. Here we will note that a numerically 
constructed curve (\textit{evolvent}) is $2^{-m}$ accurate approximation of 
the theoretical Peano curve, where $m$ is an evolvent construction parameter.

@@

The first series of experiments has been carried out on the two-dimensional problems from the classes $F_{gr}$, GKLS 
\textit{Simple}, and GKLS \textit{Hard} (100 functions from each class). Table~\ref{tab1} presents the averaged 
numbers of trials executed by GSA with the use of evolvents ($K_e$), nested optimization scheme ($K_n$), and adaptive 
nested optimization scheme ($K_{an}$). Fig.~\ref{fig1} and Fig.~\ref{fig2}(a,b) present the operating characteristics of the 
algorithms obtained on the problem classes $F_{gr}$, GKLS \textit{Simple}, and GKLS \textit{Hard} respectively. The solid 
line corresponds to the algorithm using the evolvents (GSA-E), short dashed line -- to the adaptive nested optimization scheme (GSA-AN), and 
the long dashed line -- to the nested optimization scheme (GSA-N). The results of experiments demonstrate that GSA with the use of 
the adaptive nested optimization scheme shows almost the same speed as compared to GSA with the evolvents, and 
both algorithms considerably exceed the algorithm using the nested optimization scheme. Therefore further experiments 
were limited to the comparison of different variants of the adaptive  dimensionality reduction scheme.

\begin{table}
\centering
\caption{Average number of trials for 2D problems.}\label{tab1}
\begin{tabular}{lccc}
\hline\noalign{\smallskip}
 &  $\;\;\;F_{gr}\;\;\;$  &  GKLS \textit{Simple} &  GKLS \textit{Hard} \\
\noalign{\smallskip}\hline\noalign{\smallskip}
 $K_e$ & 180  & 252 & 674 \\
 $K_n$ & 341  & 697 & 1252 \\
 $K_{an}$ & 215  & 279 & 815 \\
\noalign{\smallskip}\hline
\end{tabular}
\end{table}

@@

The second series of experiments has been carried out on the four-dimensional problems from the classes GKLS 
\textit{Simple} and GKLS \textit{Hard} (100 functions of each class). Table~\ref{tab2}  presents the averaged numbers 
of trials executed by GSA with the use of the adaptive nested optimization scheme ($K_{an}$) and block adaptive nested 
optimization scheme ($K_{ban}$) with two levels of subproblems of equal dimensionality $N_1=N_2=2$. 
Note that when solving the problem of the dimensionality $N=4$ using the initial variant of the adaptive scheme, four 
levels of one-dimensional subproblems are formed that complicates the processing of these ones.

Fig.~\ref{fig3}(a,b) presents the operating characteristics of the algorithms obtained on the classes GKLS \textit{Simple} 
and GKLS \textit{Hard} respectively. The dashed line corresponds to GSA using the adaptive nested optimization 
scheme (GSA-AN), the solid line -- the block adaptive nested optimization scheme (GSA-BAN). 
The results of experiments demonstrate the use of the block adaptive nested optimization scheme provides a considerable 
gain in the number of trials (up to 35\%) as compared to the initial adaptive nested optimization scheme.

\begin{table}
\centering
\caption{Average number of trials for 4D problems.}\label{tab2}
\begin{tabular}{lcc}
\hline\noalign{\smallskip}
 &    GKLS \textit{Simple} &  GKLS \textit{Hard} \\
\noalign{\smallskip}\hline\noalign{\smallskip}
 $K_{an}$  &  21747 & 35633 \\
 $K_{ban}$ &  13894 & 31620 \\
\noalign{\smallskip}\hline
\end{tabular}
\end{table}
